Synthesizing Quotient Lenses

### 2018-03-09 papers/QuotientLenses/main.tex
Comments on related work

\iffalse
Bidirectional transformations have been used in diverse areas of computing
where they arise as parsers and pretty printers, marshallers
and unmarshallers, serializers and deserializers, database views and view
updaters and many others\sam{TODO put citations}. Such transformations have
been extensively studied since they were proposed as a solution to the classical
{\em view-update problem} in the database community, where the challenge is to
derive a program that extracts a view of data from a source, as well as a
program that folds view updates back into the source safely and correctly.
\fi
This paper builds on the work of Foster et al~\cite{quotientlenses} who
introduced the theory of quotient lenses and implemented quotient lenses as a
refinement of the bidirectional string processing language Boomerang.
In Boomerang, the source and target types are specified using regular
expressions, and the equivalence relations are expressed using canonizers, which
are functions that map elements of a regular language to their canonical
representative. Boomerang canonizers can express a very broad class of
equivalence relations between regular languages, but actually doing so is often
difficult with complicated equivalence relations. For instance, Boomerang's
in-build permutation combinator does not account for separators in between the
regular expressions. Also, this combinator permutes regular expressions rather
than canonizers, thus making it difficult to express nested permutations that
occur in many data formats, especially XML and XML-like formats.

In their paper, Foster et al also discuss other bidirectional programming
languages that support quotienting of data including XSugar~\cite{xsugar},
biXid~\cite{bixid} and X/Inv~\cite{Hu2004,Mu2004,Mu2006}. XSugar programs
bidirectionally convert data stored in XML and ASCII formats respectively, with the
transformations specified by a pair of unambiguous grammars. The quotienting
occurs on the XML side by use of a generic canonizer that standardizes the
representation of trees. Well-formed XSugar programs are guaranteed to be
bijective modulo an equivalence relation that captures XML normalization. biXid
programs convert between pairs of XML documents, with the XML formats specified
using a pair of grammars as in XSugar. However, biXid grammars can be ambiguous.
This ambiguity is what allows biXid to express equivalences on the data.
Finally, Foster et al discuss a possible connection with the languages
X and Inv which support a primitive duplication combinator that does not work
well with the lens laws, but can be expressed using Boomerang quotient lenses.

As far as synthesis goes, though there is a good deal of recent research on
synthesizing unidirectional string
transformations~\cite{singh2012learning,le-pldi-2014,gulwani-popl-2014,perelman2014test,Singh:blinkfill},
the system Optician, which we introduced in ~\cite{optician}, is the first attempt
to synthesize bidirectional transformations that we are aware of. In that
publication, we compared Optician to two of these unidirectional string
transformers, Flash Fill~\cite{gulwani-popl-2014} and
FlashExtract~\cite{le-pldi-2014}, and found that these tools were unsuccessful
in synthesizing the complex transformatiunons that can be expressed in Optician.

While Optician improved on prior effors at synthesizing unidirectional string
transformers and introduced synthesis for bidirectional string transformations,
Optician was only able to synthesize fully bijective data transformations. Our
new system \Name{} improves on Optician in that \Name{} loosens this
restriction, and is able to synthesize data which is bijective when put in
canonical form. \Name{} is therefore a refinement of Optician since the
new synthesis algorithm (i.e. the quotient lens synthesis algorithm) restricted
to ``pure'' regular expressions degenerates to the synthesis algorithm that
we introduced in Optician. Indeed, as we mentioned in the evaluation section,
\Name{} was able to synthesize all of Optician's benchmarks at a speed
competitive with Optician, though there was additional overhead in calculating
the whole language $W(c)$ and the kernel language $K(c)$ of QREs, which
resulted in a very slight decrease in performance.

Much of the research in synthesis assumes that the synthesizer is provided with
a collection of examples. Optician and Optometrist differ in that they requires
that the programmer supplies both examples {\em and} format descriptions in the
form of regular expressions or QREs.  There is a trade-off here.  On the one
hand, a user must have some programming expertise to write regular expression
(or QRE) specifications and it requires some work. On the other hand, such
specifications provide a great deal of information to the synthesis system,
which decreases the number of examples needed (often to zero), makes the system
scale well, and allows it to handle large, complex formats.  By providing these
format specifications, the synthesis engine does not have to both infer the
format of the data as well as the transformations on it, obviating the need to
infer tricky formats like those involving nested iterations. Furthermore, by
focusing on bidirectional transformations, we limit the space of synthesized
functions to bijective ones, reducing the search space, and the expressiveness
of the search space.

>>>>

\iffalse
Bidirectional transformations have been used in diverse areas of computing
where they arise as parsers and pretty printers, marshallers
and unmarshallers, serializers and deserializers, database views and view
updaters and many others\sam{TODO put citations}. Such transformations have
been extensively studied since they were proposed as a solution to the classical
{\em view-update problem} in the database community, where the challenge is to
derive a program that extracts a view of data from a source, as well as a
program that folds view updates back into the source safely and correctly.
\fi
This paper builds on the work of Foster et al~\cite{quotientlenses} who
introduced the theory of quotient lenses and implemented quotient lenses as a
refinement of the bidirectional string processing language
Boomerang~\cite{boomerang}.  
In Boomerang, the source and target types are specified using regular
expressions, and the equivalence relations are expressed using canonizers, which
are functions that map elements of a regular language to their canonical
representative\bcp{If their use of the word is exactly what we've been using
throughout, then why define it?  Or, if we mean their exact set of
expressible canonizers, rather than the core concept of canonizer, can we be
precise about that distinction?}. Boomerang canonizers can express a very
broad class of 
equivalence relations between regular languages, but actually doing so is often
difficult with complicated equivalence relations. For instance, Boomerang's
in-build permutation combinator does not account for separators in between the
regular expressions\bcp{This sounds like a kind of trivial distinction.}. Also, this combinator permutes regular expressions rather
than canonizers, thus making it difficult to express nested permutations that
occur in many data formats, especially XML and XML-like formats.

Foster et al.\relax also discuss other bidirectional programming
languages that support quotienting of data including XSugar~\cite{xsugar},
biXid~\cite{bixid} and X/Inv~\cite{Hu2004,Mu2004,Mu2006}.\bcp{The fact that
  Fosater et al discuss these things is not germane to the present
  discussion.  We need to make our own comparisons (unless they are
  exactly the same, in which case we can just refer to Foster et al.).}
XSugar programs 
bidirectionally convert data stored in XML and ASCII formats respectively, with the
transformations specified by a pair of unambiguous grammars. The quotienting
occurs on the XML side by use of a generic canonizer that standardizes the
representation of trees. Well-formed XSugar programs are guaranteed to be
bijective modulo an equivalence relation that captures XML normalization. biXid
programs convert between pairs of XML documents, with the XML formats specified
using a pair of grammars as in XSugar. However, biXid grammars can be ambiguous.
This ambiguity is what allows biXid to express equivalences on the data.
Finally, Foster et al discuss a possible connection with the languages
X and Inv which support a primitive duplication combinator that does not work
well with the lens laws, but can be expressed using Boomerang quotient
lenses.\bcp{Same comment.}

\BCP{Are there really no bidirectional languages since 2008 that support any
form of quotienting??}

As far as synthesis goes, though there is a good deal of recent research on
synthesizing unidirectional string
transformations~\cite{singh2012learning,le-pldi-2014,gulwani-popl-2014,perelman2014test,Singh:blinkfill},
the system Optician, which we introduced in ~\cite{optician}, is the first attempt
to synthesize bidirectional transformations. In that
publication, we\bcp{confusing: Not the same ``we''!} compared Optician to
two of these unidirectional string 
transformers, Flash Fill~\cite{gulwani-popl-2014} and
FlashExtract~\cite{le-pldi-2014}, and found that these tools were unsuccessful
in synthesizing the complex transformatiunons that can be expressed in Optician.

\bcp{This paragraph seems redundant.}While Optician improved on prior effors
at synthesizing unidirectional string 
transformers and introduced synthesis for bidirectional string transformations,
Optician was only able to synthesize fully bijective data transformations. Our
new system \Name{} improves on Optician in that \Name{} loosens this
restriction, and is able to synthesize data which is bijective when put in
canonical form. \Name{} is therefore a refinement of Optician since the
new synthesis algorithm (i.e. the quotient lens synthesis algorithm) restricted
to ``pure'' regular expressions degenerates to the synthesis algorithm that
we introduced in Optician. Indeed, as we mentioned in the evaluation section,
\Name{} was able to synthesize all of Optician's benchmarks at a speed
competitive with Optician, though there was additional overhead in calculating
the whole language $W(c)$ and the kernel language $K(c)$ of QREs, which
resulted in a very slight decrease in performance.

Much of the research in synthesis assumes that the synthesizer is provided with
a collection of examples. Optician and Optometrist differ in that they requires
that the programmer supplies both examples {\em and} format descriptions in the
form of regular expressions or QREs.\bcp{But we are far from the only ones
  to consider type-based synthesis!!}  There is a trade-off here.  On the one
hand, a user must have some programming expertise to write regular expression
(or QRE) specifications, and it requires some work. On the other hand, such
specifications provide a great deal of information to the synthesis system,
which decreases the number of examples needed (often to zero), makes the system
scale well, and allows it to handle large, complex formats.  By providing these
format specifications, the synthesis engine does not have to both infer the
format of the data as well as the transformations on it, obviating the need to
infer tricky formats like those involving nested iterations. Furthermore, by
focusing on bidirectional transformations, we limit the space of synthesized
functions to bijective ones, reducing the search space, and the expressiveness
of the search space.
